---
title: "Kinship & Unemployment Benefits — Data Import & Exploration"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Packages

```{r}
# install.packages(c("tidyverse", "haven", "skimr", "remotes"))
# remotes::install_github("xmarquez/democracyData")
# install.packages("countrycode")

library(tidyverse)
library(haven)
library(skimr)
library(democracyData)
library(countrycode)
```

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

read_wb <- function(folder) {
  f <- list.files(paste0("data/", folder), pattern = "\\.csv$", full.names = TRUE)
  read_csv(f, skip = 4, show_col_types = FALSE)
}

pivot_wb <- function(df, value_name) {
  df %>%
    select(-`Indicator Name`, -`Indicator Code`) %>%
    rename(country = `Country Name`, iso3c = `Country Code`) %>%
    pivot_longer(
      cols      = matches("^[0-9]{4}$"),
      names_to  = "year",
      values_to = value_name
    ) %>%
    mutate(year = as.integer(year)) %>%
    filter(!is.na(.data[[value_name]]))
}
```

#################################################################################################################################
## Kinship Index
#################################################################################################################################

```{r}
kinship_df <- read_csv("data/kinship_df.csv", show_col_types = FALSE)

glimpse(kinship_df)
```
```{r}
summary(kinship_df)
```
```{r}
# check for NAs
sum(is.na(kinship_df$kinship_score))
```
```{r}
#  Rename isocode to iso3c for consistent merging with all other datasets later
kinship_df <- kinship_df %>% rename(iso3c = isocode)
```

flag what standardisation will look like later:
kinship_df <- kinship_df %>% mutate(kinship_std = scale(kinship_score)[,1])

The distribution is worth noting for your paper: median is 0.75 and 75th percentile is 0.92 — meaning most countries in the world have high kinship scores. Low-kinship countries (Western Europe, Scandinavia) are the minority. This is actually good for the paper — it means your zeros in the dependent variable will largely overlap with high-kinship observations, which is exactly what Enke predicts.


#################################################################################################################################
<!-- ## Unemployment Benefits Expenditure — ILO (% GDP) -->
#################################################################################################################################

```{r}
unemployment_benefits_exp_raw <- read_csv(
  "data/SDG_0131_SEX_SOC_RT_A-filtered-2026-05-25.csv",
  show_col_types = FALSE
)
glimpse(unemployment_benefits_exp_raw)
```

```{r}
summary(unemployment_benefits_exp_raw)
```

```{r}
# Check what social protection functions are available
unemployment_benefits_exp_raw %>%
  distinct(classif1.label) %>%
  print(n = 50)
```

```{r}
unemployment_benefits_exp_raw %>% distinct(sex.label)
```

```{r}
# step 1 — clean ILO as before
unemp_coverage <- read_csv(
  "data/SDG_0131_SEX_SOC_RT_A-filtered-2026-05-25.csv",
  show_col_types = FALSE
) %>%
  filter(
    sex.label      == "Total",
    classif1.label == "Function: Unemployed receiving unemployment benefits"
  ) %>%
  select(country = ref_area.label, year = time, unemp_coverage_pct = obs_value) %>%
  mutate(
    iso3c = countrycode(country, origin = "country.name", destination = "iso3c"),
    iso3c = if_else(country == "Kosovo", "XKX", iso3c)
  ) %>%
  filter(!is.na(iso3c))
```

```{r}
# step 2 — restrict to kinship countries only, THEN check years
unemp_coverage %>%
  filter(iso3c %in% kinship_df$iso3c) %>%
  group_by(year) %>%
  summarise(
    n_countries = n(),
    zero_share  = mean(unemp_coverage_pct == 0, na.rm = TRUE)
  ) %>%
  filter(year >= 2008, year <= 2019) %>%
  arrange(year)
```




```{r}
table(unemp_coverage$year)
```

```{r}
mean(unemp_coverage$unemp_coverage_pct == 0, na.rm = TRUE)

```

```{r}
unemp_coverage %>% filter(is.na(iso3c)) %>% distinct(country)

```


```{r}

unemp_coverage <- unemployment_benefits_exp_raw %>%
  filter(
    sex.label == "Total",
    classif1.label == "Function: Unemployed receiving unemployment benefits"
  ) %>%
  select(
    country            = ref_area.label,
    year               = time,
    unemp_coverage_pct = obs_value
  ) %>%
  mutate(
    iso3c = countrycode(country, origin = "country.name", destination = "iso3c"),
    iso3c = if_else(country == "Kosovo", "XKX", iso3c)
  ) %>%
  filter(!is.na(iso3c))

cat("Countries in 2019:", unemp_coverage %>% filter(year == 2019) %>% nrow(), "\n")
cat("Zero share in 2019:", mean(unemp_coverage$unemp_coverage_pct[unemp_coverage$year == 2019] == 0, na.rm = TRUE), "\n")
```
```{r}
mean(unemp_coverage$unemp_coverage_pct == 0, na.rm = TRUE)
```

#################################################################################################################################
## GDP per Capita PPP — World Bank
#################################################################################################################################

```{r}
gdp_raw <- read_csv("data/GDP per capita/API_NY.GDP.PCAP.PP.KD_DS2_en_csv_v2_1700.csv")

glimpse(gdp_raw)
```

#################################################################################################################################
## Unemployment Rate — World Bank
#################################################################################################################################

```{r}
unemployment_raw <- read_csv("data/Unemployment rate/API_SL.UEM.TOTL.ZS_DS2_en_csv_v2_115692.csv")

glimpse(unemployment_raw)
```

#################################################################################################################################
## Population Aged 65+ (%) — World Bank
#################################################################################################################################

```{r}
pop_elderly_raw <- read_csv("data/Population over 65/API_SP.POP.65UP.TO.ZS_DS2_en_csv_v2_115532.csv")

glimpse(unemployment_raw)
```


#################################################################################################################################
## Democracy Score — Polity V
#################################################################################################################################

```{r}
democracy_raw <- download_polity_annual()

glimpse(democracy_raw)
```

```{r}
# polity2 is the cleaned score: -10 (autocracy) to +10 (democracy)
# Values below -10 are transition/interruption codes, excluded here
democracy <- democracy_raw %>%
  select(iso3c = scode, country = polityIV_country, year, polity2) %>%
  filter(!is.na(polity2), polity2 >= -10)

range(democracy$year, na.rm = TRUE)
range(democracy$polity2, na.rm = TRUE)
```

## Year Selection

The kinship index is cross-sectional (one value per country), so we need to pick a single year for all panel datasets. We check how many countries have complete data across all six sources for candidate years 2015–2019.

```{r}
for (yr in 2015:2019) {
  overlap <- Reduce(intersect, list(
    unemp_benefits %>% filter(year == yr) %>% pull(iso3c),
    gdp            %>% filter(year == yr) %>% pull(iso3c),
    unemployment   %>% filter(year == yr) %>% pull(iso3c),
    pop_elderly    %>% filter(year == yr) %>% pull(iso3c),
    democracy      %>% filter(year == yr) %>% pull(iso3c),
    kinship$isocode
  ))
  cat("Year", yr, "->", length(overlap), "countries with complete data\n")
}
```
