---
title: "Kinship & Unemployment Benefits — Data Import & Exploration"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Packages

```{r}
# install.packages(c("tidyverse", "haven", "skimr", "remotes"))
# remotes::install_github("xmarquez/democracyData")

library(tidyverse)
library(haven)
library(skimr)
library(democracyData)
```

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

read_wb <- function(folder) {
  f <- list.files(paste0("data/", folder), pattern = "\\.csv$", full.names = TRUE)
  read_csv(f, skip = 4, show_col_types = FALSE)
}

pivot_wb <- function(df, value_name) {
  df %>%
    select(-`Indicator Name`, -`Indicator Code`) %>%
    rename(country = `Country Name`, iso3c = `Country Code`) %>%
    pivot_longer(
      cols      = matches("^[0-9]{4}$"),
      names_to  = "year",
      values_to = value_name
    ) %>%
    mutate(year = as.integer(year)) %>%
    filter(!is.na(.data[[value_name]]))
}
```

## Kinship Index

```{r}
kinship_df <- read_csv("data/kinship_df.csv", show_col_types = FALSE)

glimpse(kinship_df)
```


## Unemployment Benefits Expenditure — ILO (% GDP)

```{r}
unemployment_benefits_exp_raw <- read_csv(
  "data/SDG_0131_SEX_SOC_RT_A-filtered-2026-05-25.csv",
  show_col_types = FALSE
)
glimpse(unemployment_benefits_exp_raw)
```

```{r}
# Check what social protection functions are available
unemployment_benefits_exp_raw %>%
  distinct(classif1, classif1.label) %>%
  print(n = 50)
```

```{r}
unemployment_benefits_exp_raw %>%
  distinct(classif1, classif1.label) %>%
  print(n = 50)
```

```{r}
unemployment_benefits_exp_raw %>% distinct(sex, sex.label)
```

```{r}
unemp_benefits <- unemployment_benefits_exp_raw %>%
  filter(
    sex.label == "Sex: Total",
    str_detect(classif1.label, regex("unemploy", ignore_case = TRUE))
  ) %>%
  select(
    iso3c              = ref_area,
    country            = ref_area.label,
    year               = time,
    unemp_benefits_pct = obs_value
  )

table(unemp_benefits$year)
```

```{r}
# Filter to unemployment function, total sex
ilo_benefits <- ilo_raw %>%
  filter(
    sex.label == "Sex: Total",
    str_detect(classif1.label, regex("unemploy", ignore_case = TRUE))
  ) %>%
  select(
    iso3c              = ref_area,
    country_ilo        = ref_area.label,
    year               = time,
    unemp_benefits_pct = obs_value
  )

table(ilo_benefits$year)
```

```{r}
# Share of exact zeros — confirms left-censoring at 0, justifying Tobit
mean(unemp_benefits$unemp_benefits_pct == 0, na.rm = TRUE)
```


## GDP per Capita PPP — World Bank

```{r}
gdp_raw <- read_csv("data/API_NY.GDP.PCAP.PP.KD_DS2_en_csv_v2_1700.csv")

glimpse(gdp_raw)
```


## Unemployment Rate — World Bank

```{r}
unemployment <- read_wb("Unemployment rate") %>% pivot_wb("unemp_rate")
range(unemployment$year)
```


## Population Aged 65+ (%) — World Bank

```{r}
pop_elderly <- read_wb("Population over 65") %>% pivot_wb("pop_over_65")
range(pop_elderly$year)
```



## Democracy Score — Polity Vx

```{r}
democracy_raw <- download_polity_annual()

glimpse(democracy_raw)
```

```{r}
# polity2 is the cleaned score: -10 (autocracy) to +10 (democracy)
# Values below -10 are transition/interruption codes, excluded here
democracy <- democracy_raw %>%
  select(iso3c = scode, country = polityIV_country, year, polity2) %>%
  filter(!is.na(polity2), polity2 >= -10)

range(democracy$year, na.rm = TRUE)
range(democracy$polity2, na.rm = TRUE)
```

## Year Selection

The kinship index is cross-sectional (one value per country), so we need to pick a single year for all panel datasets. We check how many countries have complete data across all six sources for candidate years 2015–2019.

```{r}
for (yr in 2015:2019) {
  overlap <- Reduce(intersect, list(
    unemp_benefits %>% filter(year == yr) %>% pull(iso3c),
    gdp            %>% filter(year == yr) %>% pull(iso3c),
    unemployment   %>% filter(year == yr) %>% pull(iso3c),
    pop_elderly    %>% filter(year == yr) %>% pull(iso3c),
    democracy      %>% filter(year == yr) %>% pull(iso3c),
    kinship$isocode
  ))
  cat("Year", yr, "->", length(overlap), "countries with complete data\n")
}
```
